---
title: "PML - Assignment"
author: "lukas landzaat"
date: "11 May 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goal

The goal of your project is to predict the manner in which they did the exercise.
This is the "classe" variable in the training set.

#### Step 1 - Getting the Data

```{r echo = FALSE}
# We will use caret for all ML related tasks
library(caret)

# Setting the seed for reproducibility
setwd('~')
set.seed(12345)

# Enable Parallel Processing
library(parallel)
library(doParallel)

# Source urls
url <- c('https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv',
         'https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv')

setNames <- vector(len = length(url))
count <- 1

for (i in url) {
  setName <- strsplit(strsplit(strsplit(i, '/')[[1]][5], '\\.')[[1]][1], '-')[[1]][2]
  setNames[count] <- setName
  count <- count + 1
}

sources <- data.frame(setNames, url)
sources$url <- as.character(url)

for (i in setNames) {
  filename <- paste('pml-', i, '.csv', sep = '')
  download.file(sources[(sources$setNames == i), 2], destfile = paste('~/', filename, sep = ''), method = "curl")
}
```

#### 3 - Cleaning the Data

```{r}
setwd('~')

# Trainset

# Here we treat values that contain 'NA' or are blank ("") as missing
trainSource <- read.csv('pml-training.csv', na.strings = c("NA", ""))

# We will do a one-to-many logistic regression later on
# Therefore we add 5 indicator columns [A:E]
for (i in levels(trainSource$classe)) {
  columnName <- paste('classe', i, sep = '')
  trainSource[(trainSource$classe == i), columnName] <- 1
  trainSource[is.na(trainSource[,columnName]), columnName] <- 0
  trainSource[,columnName] <- as.factor(trainSource[,columnName])
}

# Only selecting relevant columns
logicalColumns <- grepl('.*arm.*|.*dumbbell.*|.*belt.*|.*forearm.*|.*classe.*', colnames(trainSource))
trainSource <- trainSource[ , logicalColumns]

# Check for % of missing values..!
columnsNA <- vector(length = dim(trainSource)[2])
count <- 1

for (i in names(trainSource)) {
  columnsNA[count] <- sum(is.na(trainSource[,i])) / length(trainSource[,i])
  count <- count + 1
}

# Going to only include columns where there are 0 NA values
columnsNaLogical <- columnsNA == 0
trainSource <- trainSource[ , columnsNaLogical]

## TestSet
testSource <- read.csv('pml-testing.csv', na.strings = c("NA", ""))

# Only selecting relevant columns
logicalColumns <- grepl('.*arm.*|.*dumbbell.*|.*belt.*|.*forearm.*|.*classe.*', colnames(testSource))
testSource <- testSource[ , logicalColumns]
testSource <- testSource[ , columnsNaLogical[1:152]]

```

#### 2 - Dividing the training data into ValidationSet, TrainSet & TestSet

```{r}
# First we separate out a ValidationSet set from trainSource
inValidation <- createDataPartition(y = trainSource$classe, p = 0.3, list = FALSE)
ValidationSet <- trainSource[inValidation,] # 30% of Data
NonValidationSet <- trainSource[-inValidation,] # 70% of Data

# Second we sub-divide the NonValidationSet into a TrainSet & TestSet
inTrain <- createDataPartition(y = NonValidationSet$classe, p = 0.6, list = FALSE)
TrainSet <- NonValidationSet[inTrain,] # 60% of 70%
TestSet <- NonValidationSet[-inTrain,] # 40% of 70%

# Separating numeric variables from factor variables.
logicalNumerics <- vector(length = dim(trainSource)[2])
count <- 1

for (i in names(trainSource)) {
  if (class(trainSource[,i]) %in% c('numeric', 'integer') == TRUE) { logicalNumerics[count] <- TRUE }
  count <- count + 1
}

classe <- names(TrainSet) == 'classe'
logicalNumerics <- (logicalNumerics + classe) > 0
logicalFactors <- !logicalNumerics

# Removing redundant variables
rm(classe, columnName, count, i)
```

#### 3 - Training Models

###### 3.1 Training Model 1 - Classification Tree

```{r echo = FALSE}
cluster <- makeCluster(detectCores() - 1) # convention to leave 1 core for OS
registerDoParallel(cluster)
fitControl <- trainControl(method = "cv", number = 10, allowParallel = TRUE)
modelRPART <- train(classe ~., 
                    method= 'rpart', 
                    data = TrainSet[ , !names(TrainSet) %in% c('classeA', 'classeB', 'classeC', 'classeD', 'classeE')], 
                    trControl = fitControl)
stopCluster(cluster)
```

###### 3.2 Training Model 2 - RandomForest

```{r echo = FALSE}
# Random Forest with additional CV
cluster <- makeCluster(detectCores() - 1) # convention to leave 1 core for OS
registerDoParallel(cluster)
fitControl <- trainControl(method = "cv", 
                           number = 10, 
                           allowParallel = TRUE)
modelRF <- train(classe ~., method= 'rf', 
                 data = TrainSet[ , !names(TrainSet) %in% c('classeA', 'classeB', 'classeC', 'classeD', 'classeE')], 
                 trControl = fitControl)
stopCluster(cluster)
```

###### 3.3 Training Model 3 - One-to-Many Logistic Regression

```{r echo = FALSE}
cluster <- makeCluster(detectCores() - 1) # convention to leave 1 core for OS
registerDoParallel(cluster)
fitControl <- trainControl(method = "cv", 
                           number = 10, 
                           allowParallel = TRUE)
modelGLMclasseA <- train(classeA ~., 
                         method = 'glm', 
                         data = TrainSet[ , !names(TrainSet) %in% c('classe', 'classeB', 'classeC', 'classeD', 'classeE')], 
                         trControl = fitControl, 
                         family = binomial(link = "logit"))
modelGLMclasseB <- train(classeB ~., 
                         method = 'glm', 
                         data = TrainSet[ , !names(TrainSet) %in% c('classe', 'classeA', 'classeC', 'classeD', 'classeE')], 
                         trControl = fitControl, 
                         family = binomial(link = "logit"))
modelGLMclasseC <- train(classeC ~., 
                         method = 'glm', 
                         data = TrainSet[ , !names(TrainSet) %in% c('classe', 'classeA', 'classeB', 'classeD', 'classeE')], 
                         trControl = fitControl, 
                         family = binomial(link = "logit"))
modelGLMclasseD <- train(classeD ~., 
                         method = 'glm', 
                         data = TrainSet[ , !names(TrainSet) %in% c('classe', 'classeA', 'classeB', 'classeC', 'classeE')], 
                         trControl = fitControl, 
                         family = binomial(link = "logit"))
modelGLMclasseE <- train(classeE ~., 
                         method = 'glm', 
                         data = TrainSet[ , !names(TrainSet) %in% c('classe', 'classeA', 'classeB', 'classeC', 'classeD')], 
                         trControl = fitControl, 
                         family = binomial(link = "logit"))
stopCluster(cluster)
```

#### 4 - Model Accuracy 

###### 4.1 Results: Model 1 - Classification Tree

```{r}
# Evaluate on TrainSet
modelRPART[[4]][1,2]

# Evaluate on TestSet
confusionMatrix(TestSet$classe, 
                predict(modelRPART, 
                TestSet[ , !names(TrainSet) %in% c('classeA', 'classeB', 'classeC', 'classeD', 'classeE')]))[[3]][1]
```

###### 4.2 Results: Model 2 - RandomForest

```{r}
# Evaluate on TrainSet
modelRF[[4]][1,2]

# Evaluate on TestSet
confusionMatrix(TestSet$classe, 
                predict(modelRF, 
                TestSet[ , !names(TrainSet) %in% c('classeA', 'classeB', 'classeC', 'classeD', 'classeE')]))[[3]][1]
```

###### 4.3 Results: Model 3 - One-to-Many Logistic Regression

```{r}
# First we get the predictions of all our models
A <- predict(modelGLMclasseA, 
             TestSet[ , !names(TrainSet) %in% c('classe', 'classeB', 'classeC', 'classeD', 'classeE')], 
             type = 'prob')[,2]
B <- predict(modelGLMclasseB, 
             TestSet[ , !names(TrainSet) %in% c('classe', 'classeA', 'classeC', 'classeD', 'classeE')], 
             type = 'prob')[,2]
C <- predict(modelGLMclasseC, 
             TestSet[ , !names(TrainSet) %in% c('classe', 'classeA', 'classeB', 'classeD', 'classeE')], 
             type = 'prob')[,2]
D <- predict(modelGLMclasseD, 
             TestSet[ , !names(TrainSet) %in% c('classe', 'classeA', 'classeB', 'classeC', 'classeE')], 
             type = 'prob')[,2]
E <- predict(modelGLMclasseE, 
             TestSet[ , !names(TrainSet) %in% c('classe', 'classeA', 'classeB', 'classeC', 'classeD')], 
             type = 'prob')[,2]
predictions <- data.frame(A, B, C, D, E)

# We will now determine for each instance which model has the highest prediction
# And assign the values the corresponding class
maxValues <- apply(predictions, 1, max)
predictions$final <- NULL

for (i in 1:dim(predictions)[1]) {
  for (j in names(predictions)[1:5]) {
    if (predictions[i,j] == maxValues[i]) {
      predictions[i,'final'] <- j
      break
    }
    else { next }
  }
}

# This looks like this:
head(predictions)

# We can now evaluate the final accuracy:
confusionMatrix(TestSet$classe, predictions$final)[[3]][1]
```

#### 5 - Choosing Optimal Model & Assessing Accuracy on ValidationSet

It seems evident that the **RandomForest** algorithm has by far the highest accuracy.

However, we used the TestSet to evaluate which algorithm works best. Therefore we must use the ValidationSet
to see what the final estimated out of sample accuracy is of our chosen algorithm:

```{r}
# Evaluate on ValidationSet
confusionMatrix(ValidationSet$classe, 
                predict(modelRF, 
                ValidationSet[ , !names(TrainSet) %in% c('classeA', 'classeB', 'classeC', 'classeD', 'classeE')]))[[3]][1]
```

This is the final out of sample accuracy!

#### 6 - Predicting classe of the 20 Test Cases

```{r}
# Aggregate Results:
table(predict(modelRF, testSource))

# Results by row:
prediction <- predict(modelRF, testSource)
row <- 1:20
data.frame(row, prediction)
```
